Context-dependent consequences of lagged effects in demographic models

### 2025-04-23 docs/ipm_comparison_ms.qmd
updating plots and ms

We have previously shown that the effects of precipitation extremes on the demographic vital rates of an Amazonian understory herb (*Heliconia acuminata*, Heliconiaceae) can be delayed up to 36 months [@scottDelayedEffectsClimate2022], with the presence and duration of these lagged effects varying by vital rate and habitat. We parameterized three classes of Integral Projection Models - a deterministic IPM, a stochastic IPM, and a stochastic IPM with lagged effects of precipitation on vital rates - for populations in two habitat type (i.e., continuous forest vs. forest fragments). Based on previous studies [@bruna_are_2003; @bruna_demographic_2005; @bruna_experimental_2002; @bruna_habitat_2002] and demographic theory [@caswell2001; @tuljapurkar_population_1990] we predicted that:\

(i) projections of $\lambda$ would be lower in forest fragments than in continuous forest for all model types,
(ii) projections of $\lambda$ from deterministic models would be higher than those of stochastic models,
(iii) projections of $\lambda$ would be lowest for models with lagged effects, and
(iv) populations would be more skewed towards pre-reproductive size classes in fragments that forest, regardless of model structure

@@

A complete description of the demographic methods, data, and analyses to date can be found in @bruna_demography_2023. Briefly, in 1997--1998 a series of 5000 m^2^ plots were established in the BDFFP reserves: N=6 in Continuous Forest and N=4 in 1-ha Forest Fragments. All of the *Heliconia acuminata* in these plots were marked and measured; the plots were censused annually, at which time a team recorded the size of surviving individuals, marked and measured new seedlings, and identified any previously marked plants that died. Each plot was also surveyed 4-5 times during the flowering season to identify reproductive plants; in our site *H. acuminata* begin flowering early in the rainy season (e.g., January) and most reproductive plants produce a single infloresence (range = 1--7) with 20--25 flowers [@bruna_habitat_2002]. Fruits mature April-May and have 1--3 seeds per fruit ($\bar{x}=2$) that are dispersed by a thrush and several species of manakin [@uriarte_disentangling_2011]. Dispersed seeds germinate approximately 6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@bruna_habitat_2002; @bruna_seed_1999]. On average olots in CF also had more than twice as many plants as the plots in 1-ha fragments (CF median = 788, range = (201-1549); 1-ha median = 339, range = (297-400)).\

@@

We compared three classes of IPMs - Deterministic, Stochastic, and Stochastic with lagged effects of precipitation - each of which required different functional forms of their underlying vital rates models (@fig-lifecycle). All of the models were density-independent.

**(1) Deterministic IPM:** In this model, which served as the foundation for more complex models, the number and size of mature plants in year $t+1$ is determined by sub-kernels describing the survival and growth (or regression) of mature plants from year $t$ to $t+1$, the number of seedlings establishing in year $t$ that survive to $t+1$ (@eq-mature), and the number of new seedlings that enter the population (@eq-sdlg).

@@

In @eq-mature the sub-kernel $P(z^\prime, z)$ describes the size-dependent survival and growth or regression ($s(z)$ and $G(z^\prime, z)$, respectively) of mature plants from one year to the next ([@eq-P]). Sub-kernel $R(z^\prime)$ describes the survival and growth of seedlings that established in year $t$ to year $t+1$ ([@eq-R]); note that these values are not size-dependent - in all our IPMs 'seedlings' are a discrete, size-independent category with a probability $s_s$ of survival for one year following germination and establishment, with the size at which they transition into the 'mature' plant category given by $G(z^\prime, z)$. IPMs can include transitions between continuous and discrete states [@ellnerDatadrivenModellingStructured2016]); we treat seedlings as a distinct category because they have lower survival and growth in their first year than comparably sized plants [@bruna_are_2003; @scottDelayedEffectsClimate2022].

@@

@@

The number of new seedlings entering the population, $n_s(t+1)$ ([@eq-sdlg]), is described by the fecundity sub-kernel ([@eq-F]). Both the probability that a mature plant will flower, $p_f(z)$, and the number of seeds they produce, $f(z)$, are size dependent. These seeds germinate and establish as seedlings with probability $g$.

$$
F(z) = p_f(z)f(z)g
$$ {#eq-F}

Vital rate models for growth ($G_s(z^\prime)$ and $G(z^\prime, z)$), survival ($s_s$ and $s(z)$), and flowering ($p_f(z)$) were fit using the long-term demographic data [@bruna_are_2003]. For established plants, these three vital rates were modeled as a smooth function of size in the previous census using generalized additive models (GAMs) fit with the `mgcv` package [@mgcv] in `r version$version.string` [@rcoreteam2020]<!--# probably include a supplement with detail on things like basis, number of knots, etc. -->. For consistency, seedling survival and growth were also modeled using GAMs, but without size in the previous census as a predictor (i.e. intercept only models). For growth models ($G_s(z^\prime)$ and $G(z^\prime, z)$) a scaled t family distribution provided a better fit to the data than a gaussian fit as the residuals were leptokurtic with a simple Gaussian model. Size-specific fecundity was based on surveys of fruits per flowering plant [@brunaLeafNumberLeaf2021] and seeds per fruit [@brunaDatasetHeliconiaAcuminata2014] (together $f(z)$) and experimentally derived estimates of germination and establishment in forest and fragments [@bruna_effects_2002; @bruna_seed_1999] <!--# Emilio: add more info here -->.\

**(2) Stochastic IPM:** To build the stochastic, kernel-resampled IPMs we included environmental stochasticity in all vital rate models built using the long term demographic datset by adding a random effect of year (@fig-lifecycle). The random effect of year was included using a factor--smooth interaction which allowed the relationship between plant size and vital rates to vary in functional form among transition years. The kernel-resampling approach is to generate kernels corresponding to each transition year in the demographic dataset using the random smooths for year, and to iterate the IPM by drawing from these randomly. This is equivalent to the matrix selection approach for matrix population models described by @caswell2001.\

**(3) Stochastic IPM with lagged effects of precipitation on vital rates:** In our The stochastic, parameter-resampled IPMs the impacts of precipitation extremes on vital rates were modeled explicitly (*sensu* @metcalfStatisticalModellingAnnual2015). We calculated the standardized precipitation evapotranspiraton index (SPEI) for our site using a published gridded dataset based on ground measurements [@xavierDailyGriddedMeteorological2016] as described in @scottDelayedEffectsClimate2022. For all vital rate models fit using the long term demographic dataset, we modeled delayed effects of SPEI using distributed lag non-linear models with a maximum lag of 36 months [@scottDelayedEffectsClimate2022] (@fig-lifecycle). To iterate these parameter-resampled IPMs, a random sequence of SPEI values was created by sampling years of the observed monthly SPEI data. Then, 36 month lags are calculated for each year starting in February (the month of the demographic census). These values are then used to predict fitted values from the vital rates models, generating different kernels at each iteration of the IPM. With this method, the kernels of successive iterations are not entirely independent because the SPEI values used in calculating vital rates include values used in the previous two iterations, but they are ergodic.\

@@

All IPMs were constructed and iterated <!--# optimized in notes/optimizing-meshpoints.Rmd which could be turned into supplemental -->using the `ipmr` package in R [@levinIpmrFlexibleImplementation2021]. The IPMs used 100 meshpoints and the midpoint rule for calculating kernels <!--# citation? -->. For each type of IPM we iterated the model for 1000 time steps, discarding the first 100 time steps to omit transient effects. Stochastic growth rates ($\lambda_s$) were calculated as the average $ln(\lambda)$ from each time step [@caswell2001] and back-transformed to be on the same scale as deterministic lambdas for comparison. We used the distribution of established plant sizes and proportion of seedlings from the full dataset as a starting population vector. While other starting population vectors were possible, the choice is of little importance as it will only impact transient dynamics, which we aren't interested in for this study.\

To estimate uncertainty around the per-capita growth rates (lambdas), we created 500 bootstraps of the demographic dataset by sampling individual plants with replacement within each habitat. For each bootstrap, we then re-fit vital rates models (all except germination and establishment rate, fruits per flowering plant, and seeds per fruit, which were estimated using different datasets), constructed IPMs, and calculated a value for lambda as described above. We then used these bootstraped estimates of lambda to calculate bias corrected 95% confidence intervals [@ellnerDatadrivenModellingStructured2016].\

#TODO: need to add something about the comparison of population structure (figure)

This workflow was managed using the `targets` R package [@targets] which also allowed us to track computational time spent on each IPM for comparison.\

@@

Comparison of CF vs FF

@@

@@

1.  For all vital rates estimated using the long term demographic dataset, the DLNM model fit the best (dAIC = 0) followed by the model with a random effect of year, followed by the deterministic model (@tbl-aic).\
    <!--# No longer true if random effect = different smooth for each year -->

2.  Population growth rates in CF were significantly higher than in forest fragments for all IPM types (@tbl-lambdas).\

3.  Deterministic and stocuhastic-kernal similar means.

4.  These results are consistent with those of @kayeEffectStochasticTechnique2003, who found who found that the method effected stochastic lambda but relative ranking of populations was consistent.

5.  BUT lag dropped lambdas considerably. In fact, took both well below zero. This is a big, big difference. If you are doing conservation and management, a growth rate lower by 5-6 % would really rock your world.

### Population Structure

6.  @fig-pop-states has some interesting things in it:

> -   For the deterministic IPM (and the kernel-resampled IPM?) there are slightly more of the smallest plants and the largest plants in CF compared to FF (i.e. more medium sized plants in FF).

> -   For the kernel-resampled IPM (random effect of year), the fluctuations are extremely similar between CF and FF

> -   For the parameter-resampled IPM (DLNM) the size structure of the population is a LOT more variable in FF. This makes sense as we know lagged effects are more important in fragments.

> -   Also, the fluctuations in size structure in CF do not match the fluctuations in FF as well (can see this by the increased spread of points in @fig-pop-states (B)

> -   Also, in the parameter-resampled IPM (and only in this one), we see a shift toward smaller plants in FF compared to CF

### Time and Effort

4.  DLNM models take much, much longer to iterate: while the Deterministic and Kernel-resampled stochastic models took \~`r det_time` and \~`r sker_time` min to iterate (respectively), the Parameter-resampled stochastic models with lagged effects took \~`r spar_time` min.\

5.  The greater use of computational resources is likely a result of `predict()` being much slower for GAMs with 2D smooths because the number of knots is much higher compared to the GAMs used for the vital rates models in the determinsitic and kernel-resampled IPMs.\

@@

We thank **,** \_, \_\_\_ and \_\_\_ anonymous reviewers for helpful discussions and comments on the manuscript. We thank Sam Levin for his help with the `ipmr` package. Financial support was provided by the U.S. National Science Foundation (awards \_\_\_\_, and \_\_\_\_). This article is publication no. -- -- in the BDFFP Technical series. The authors declare no conflicts of interest.

@@

ERS contributed to the conceptualization, methodology, formal analysis, and led the writing of the original draft. EMB contributed to the conceptualization, methodology, and writing and also acquired funding.

@@

@@

```{r aic, results='asis'}
#| label: tbl-aic
#| tbl-cap: "Comparison of vital rate models used to build IPM.  The 'Effect of Environment' column describes how environmental effects were included in models. Those with 'none' were used to build deterministic IPMs; those with a random effect of year were used to build stochastic, kernel-resampled IPMs; and those with a distributed lag non-linear model (DLNM) were used to build stochastic, parameter-resampled IPMs. 'edf' is the estimated degrees of freedom of the penalized GAM. \u0394AIC is calculated within each habitat and vital rate combination.  \u0394AIC within 2 indicates models are equivalent."
#| 
#TODO: only show first instance of habitat and vital rate columns, not on every row.  Alternatively do something fancy with flextable or gt.

aic_tbl_df %>%
  group_by(habitat, model) %>% 
  # mutate(rank = row_number()) %>% 
  ungroup() %>% 
  mutate(habitat = toupper(habitat),
         env_effect_incl = ifelse(
           env_effect_incl != "DLNM",
           to_sentence_case(as.character(env_effect_incl)),
           as.character(env_effect_incl)
         )
  )%>% 
  mutate(
    model = fct_relevel(
      model,
      c("surv", "size", "flwr", "surv_sdlg", "size_sdlg")
    ),
    model = fct_recode(
      model,
      "Survival"          = "surv",
      "Growth"            = "size",
      "Flowering"         = "flwr",
      "Seedling survival" = "surv_sdlg",
      "Seedling growth"   = "size_sdlg"
    )
  ) %>%

@@

  arrange(habitat, model) %>% 
  select(
    "Habitat" = "habitat",
    "Vital Rate" = "model",
    "Effect of Environment" = "env_effect_incl",
    #"Model" = "RHS",
    "edf" = "df",
    "\u0394AIC" = "dAIC"
  ) %>%

pandoc.table(
  justify = c( "left", "left","left", "right",  "right"),
  full_width = FALSE
  # ,
  # caption = Comparison of vital rate models used to build IPM.  The 'Effect of Environment' column describes how environmental effects were included in models. Those with 'none' were used to build deterministic IPMs; those with a random effect of year were used to build stochastic, kernel-resampled IPMs; and those with a distributed lag non-linear model (DLNM) were used to build stochastic, parameter-resampled IPMs. 'edf' is the estimated degrees of freedom of the penalized GAM. \u0394AIC is calculated within each habitat and vital rate combination.  \u0394AIC within 2 indicates models are equivalent."
)
```

@@

lambda_table %>% 
  mutate(habitat = toupper(habitat)) %>% 
  mutate(across(c(est, lower, upper), ~format(round(.,4)))) %>% 
  mutate(lambda = glue::glue("{est} ({lower}, {upper})")) %>%
  mutate(ipm = str_replace_all(
    ipm,
    c(
      "det" = "Deterministic",
      "stoch" = "Stochastic, kernel-resampled",
      "DLNM" = "Stochastic, parameter-resampled"
    )
  )) %>% 
  select(IPM = ipm, Habitat = habitat, "$\\lambda$" = lambda) %>% 
pandoc.table(
  digits = 5,
  keep.trailing.zeros = TRUE,
  justify = c("left", "left", "right"),
  full_width = FALSE
  # ,
  # caption = "(\\#tab-lambdas) Population growth rates for continuous forest (CF) and forest fragments (FF) under different kinds of IPMs with bootstrapped, bias-corrected, 95% confidence intervals."
)

@@

```{r #fig-lifecycle,  echo=FALSE, out.width="85%",fig.align='center'}
#| fig.cap = "Lifecycle diagram of *Heliconia acuminata*. Each transition is associated with an equation for a vital rate function.  The functions shown on the diagram correspond to those used to construct a general, density-independent, deterministic IPM. The table below shows the equivalent equations for stochastic, kernel-resampled IPMs and stochastic, parameter-resampled IPMs."

```{r #fig-pop-states,  echo=FALSE, out.width="85%",fig.align='center'}

>>>>

We have previously shown that the effects of precipitation extremes on the demographic vital rates of an Amazonian understory herb (*Heliconia acuminata*, Heliconiaceae) can be delayed up to 36 months [@scottDelayedEffectsClimate2022], with the presence and duration of these lagged effects varying by vital rate and habitat. We parameterized three classes of Integral Projection Models - a deterministic IPM, a stochastic IPM, and a stochastic IPM with lagged effects of precipitation on vital rates - for populations in two habitat type (i.e., continuous forest vs. forest fragments). Based on previous studies [@bruna_are_2003; @bruna_demographic_2005; @bruna_experimental_2002; @bruna_habitat_2002] and demographic theory [@caswell2001; @tuljapurkar_population_1990] we predicted that: (i) projections of $\lambda$ from deterministic models would be higher than those of stochastic models, and that (ii) for all IPMs the projections of $\lambda$ forest fragments would be lower than for continuous forest, but that the difference would be proportionately greater for IPM with lagged effects.\

@@

A complete description of the demographic methods, data, and analyses to date can be found in @bruna_demography_2023. Briefly, in 1997--1998 a series of 5000 m^2^ plots were established in the BDFFP reserves: N=6 in Continuous Forest and N=4 in 1-ha Forest Fragments. All of the *Heliconia acuminata* in these plots were marked and measured; the plots were censused annually, at which time a team recorded the size of surviving individuals, marked and measured new seedlings, and identified any previously marked plants that died. Each plot was also surveyed 4-5 times during the flowering season to identify reproductive plants; in our site *H. acuminata* begin flowering early in the rainy season (e.g., January) and most reproductive plants produce a single infloresence (range = 1--7) with 20--25 flowers [@bruna_habitat_2002]. Fruits mature April-May and have 1--3 seeds per fruit ($\bar{x}=2$) that are dispersed by a thrush and several species of manakin [@uriarte_disentangling_2011]. Dispersed seeds germinate approximately 6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@bruna_habitat_2002; @bruna_seed_1999]. On average olots in CF also had more than twice as many plants as the plots in 1-ha fragments (CF median = 788, range = (201-1549); 1-ha median = 339, range = (297-400)).

@@

We projected the growth rate and structure of *Heliconia acuminata* populations in Continuous Forest and Forest Fragments with three classes of IPMs - Deterministic, Stochastic, and Stochastic with Lagged Environmental Effects. Each of these IPMs required different functional forms of the underlying vital rate functions used to describe the *H. acuminata* life cycle (@fig-lifecycle). All models were density-independent, with the deterministic model serving as the foundation for the more complex models.\

**(1) Deterministic IPM:** In this model the size and structure of a population in year $t+1$ is determined by the survival and growth of plants alive in year $t$ (@eq-mature) plus the number of new seedlings that entered the population (@eq-sdlg).

@@

<!-- The number and size of mature plants in year $t+1$ is determined by sub-kernels describing the survival and growth (or regression) of mature plants from year $t$ to $t+1$, the number of seedlings establishing in year $t$ that survive to $t+1$ (@eq-mature), and the number of new seedlings that enter the population (@eq-sdlg).\ -->

@eq-mature has two components. The first is the sub-kernel $P(z^\prime, z)$, which describes the size-dependent survival and growth/regression of mature plants (@eq-P):

@@

The second is sub-kernel $R(z^\prime)$, which describes the survival of seedlings established in year $t$ and their size when entering the mature plant population in year $t+1$ (@eq-R):

@@

Note that in @eq-R both the probability that new seedlings survive their first year, $s_s$, and their size at the end of this year, $G(z^\prime, z)$, are size-independent. IPMs can include transitions between individuals from a discrete state to a continuous one [i.e., from 'seedling' to 'mature plant of size $z'$', @ellnerDatadrivenModellingStructured2016]; we treat seedlings as a distinct and discrete category because they have lower survival and growth in their first year than comparably sized plants [@bruna_are_2003; @scottDelayedEffectsClimate2022].\

The number of new seedlings entering the population in year $t+1$ is a function of the number of mature plants in year $t$ and a sub-kernel describing the size-dependent fecundity of these individuals (@eq-F): $$
F(z) = p_f(z)f(z)g
$$ {#eq-F}

Both the probability that a mature plant will flower, $p_f(z)$, and the number of seeds a flowering plant will produce, $f(z)$, are size-dependent. All seeds germinate and establish as seedlings with probability $g$.\
<!-- `r version$version.string`  -->

We used the annual census data [@bruna_are_2003] to fit the deterministic vital rate functions for growth, survival, and flowering in each habitat type (i.e., Fragments, Continuous Forest; the data from all plots within a habitat class were pooled to create a single 'summary population' [@bruna_are_2003]. For established plants these were modeled as a smooth function of size in the previous census with generalized additive models (GAMs) fit with the `mgcv` library [@mgcv] for the R statistical programming language [@rcoreteam2020].<!--`r version$version.string`---> <!--# probably include a supplement with detail on things like basis, number of knots, etc. --> For consistency, seedling survival and growth were also modeled using GAMs, but without size in the previous census as a predictor (i.e. 'intercept-only' models). For growth models a scaled t family distribution provided a better fit to the data than a Gaussian fit, as the residuals with a simple Gaussian model were Leptokurtic. To model size-specific fecundity we used data on the number of fruits per flowering plant [@brunaLeafNumberLeaf2021] and seeds per fruit [@brunaDatasetHeliconiaAcuminata2014] that were collected outside of the demographic plots (to avoid altering within-plot recruitment) and experimentally-derived estimates of seed germination and seedling establishment [@bruna_effects_2002; @bruna_seed_1999] <!--# Emilio: add more info here -->.\

**(2) Stochastic IPMs:** To include temporal stochasticity in our IPMs we included a random effect of year. This was done using a factor--smooth interaction that allowed the functional form of the relationship between plant size and vital rates to vary among transition years (@fig-lifecycle). We generated kernels for every transition year using the long-term survey data [@bruna_demography_2023] and random smooths for year. We then randomly selected one of these sets of kernels to use in each iteration of the IPM. This procedure is equivalent to *'kernel resampling'* [*sensu* @metcalfStatisticalModellingAnnual2015] or matrix selection for matrix population models [@caswell2001].\

**(3) Stochastic IPMs with lagged effects of precipitation on vital rates:** We explicit modeled the lagged effects of precipitation extremes on vital rates using the procedure described in @scottDelayedEffectsClimate2022. Briefly, we first calculated the Standardized Precipitation Evapotranspiraton Index (i.e., SPEI) for our study site using a published gridded dataset based on ground measurements [@xavierDailyGriddedMeteorological2016]. After we fit vital rate models using the long-term survey data (@fig-lifecycle), we modeled delayed effects of SPEI with Distributed Lag Non-linear Models (i.e., DLNMs) with a maximum lag of 36 months [@scottDelayedEffectsClimate2022]. To iterate these parameter-resampled IPMs [*sensu* @metcalfStatisticalModellingAnnual2015] we first created a random sequence of SPEI values by sampling years of the observed (monthly) SPEI data. For every year we then calculated a lag of 36 months from the month in which that year's census was completed. These values were then used to predict the fitted values from vital rate models, which generated different kernels for each iteration of the IPM. The kernels of successive iterations are not entirely independent -- the SPEI values used to calculate vital rates include values used in the previous two iterations -- they are ergodic.\

@@

All IPMs were constructed and iterated <!--# optimized in notes/optimizing-meshpoints.Rmd which could be turned into supplemental -->using the `ipmr` package [@levinIpmrFlexibleImplementation2021] for the R statistical programming language [@rcoreteam2020]. The IPMs used 100 meshpoints and the midpoint rule for calculating kernels <!--# citation? -->. For each type of IPM we iterated the model for 1000 time steps, but discarded the first 100 time steps to omit transient effects. Stochastic growth rates ($\lambda_s$) were calculated as the average $ln(\lambda)$ from each time step [@caswell2001] and then back-transformed to allow for direct comparison with projections of $\lambda$ from deterministic models. The initial starting vector primarily influences a population's transient dynamics; we therefore used the distribution of established plant sizes and proportion of seedlings in the full demographic data set as the initial population vector for all models. Finally, we estimated the 95% confidence intervals for each IPMs projections of $\lambda$. To do so we first created 500 populations for each habitat type by sampling individual plants with replacement (i.e., bootstrapping) until the population size of each matched that of the initial population vector. We then re-fit the vital rate models for growth, survival, and flowering for each of these bootstrapped population and constructed new IPMs for each population as above (the models for germination and establishment rate, fruits per flowering plant, and seeds per fruit were not refit because these were estimated using different data sets). The projections of $\lambda$ for the new populations were then used to estimate the upper and lower 95% bias-corrected percentile intervals [@manly2018randomization; @caswell2001]. This workflow was managed using the `targets` R package [@targets], which also allowed us to track computational time spent processing and analyzing each class of IPM.\

@@

##### Comparison of CF vs FF

@@

-   Comparison of pop structure in figure (test for increasing variance with variance / F or permutation test). Villase∼nor, J. A., & González-Estrada, E. (2024). A nonparametric test for homogeneity of variances. Communications in Statistics - Theory and Methods, 54(2), 646–654. https://doi.org/10.1080/03610926.2024.2316274 OR https://www.datanovia.com/en/lessons/homogeneity-of-variance-test-in-r/#google_vignette

@@

### Lambda Comparison

1.  We found that including Delayed Life-History Events in Intergral Projection Models led to significantly lower projections of $\lambda$
2.  It is reassuring that all IPM models projected higher values of $\lambda$ in continuous forest than forest fragments (@tbl-lambdas); this is consistent with the results of previous studies comparing the relative rankings of projections from stochastic and deterministic models [e.g., @kayeEffectStochasticTechnique2003.]
3.  However, the underestimating $\lambda$ by 2-3% difference in $\lambda$ resulting projections from the different IPMs could have major implications for many of the contexts in which IPMs are brought to bear.
4.  If you are doing conservation and management, a growth rate lower by 5-6 % would really rock your world.
5.  For instance, underestimating $\lambda$ by failing to include lagged effects could be especially severe in studies of climate change, given many DLHE are climate driven.\
    <!-- CF: deterministic is > than 70% of dnlm --> <!-- FF: deterministic is > than 76% of dnlm -->

### Pop Structure Comparison

1.  Including lagged effects in IPMs also resulted in more variable projections of future population structure (@fig-pop-states a), especially in forest fragments where we have previously shown lagged effects have very large impacts on growth and survival [@scottDelayedEffectsClimate2022].
2.  Consequently, projections of population structure were also far less less consistent across habitats (@fig-pop-states b).
3.  Finally, the choice of IPM also resulted in some subtle but notable differences in projected population structure.
4.  For example, the deterministic IPM projected the population in Continuous Forest had slightly more of the smallest and largest plants than the population in Forest Fragments.
5.  In other words, populations in forest fragments had proportionatly less recruitment and fewer individuals growing into the larger and reproductivce size classes [@bruna_demographic_2005].
6.  The shift towards intermediately sized, pre-reproductive plants in forest fragments is even more dramatic when using the parameter-resampled stochastic IPM (@fig-pop-states b).\

### So...should we care?

1.  Including delayed life-history events in demographic models has important implications both projections of population growth rate and population structure.
2.  These projections may be more accurrate - the underlying vital rate models for IPMS with Lagged effects all had the best fit to the survey data (dAIC = 0, @tbl-aic). <!--# No longer true if random effect = different smooth for each year -->
3.  While this may not be critical in situations where relative differences allow for hypothesis testing, it may be especially important in conservation or manaagement settings.
4.  Including lagged effects comes at a cost, however.
5.  While Deterministic and Kernel-resampled Stochastic IPMs took only \~`r det_time` and \~`r sker_time` min to iterate (respectively), the Stochastic IPMs with using parameter-resampled kernels and lagged effects took \~`r spar_time` min.
6.  This is largely due to the required computational resources and algorithms (i.e., `predict()`), which are much slower forGeneral Additive Models (i.e., GAMs) with 2D smooths because of the much higher number of knots than for the GAMs used in Determinsitic and Kernel-resampled IPMs.
7.  Advances in computational power and access to high-performance computing resources could lower this cost.\
8.  However, computational power cannot compensate for limited data.
9.  Detecting lagged effects and evaluating their consequences requires long-term demographic data - data that are only available for a relatively small number of species, few of which are in the tropics.
10. The increasing evidence that lagged effects are ubiquitous, and that they can have major demographic impacts, underscores the need to support the collection of such long-term data, the complementary development of experimental and statistical approaches to disentangling lagged effects, and community driven efforts to identify priority or model systems for in-depth investigation.\

@@

We thank \_\_\_, \_\_\_, and \_\_\_ anonymous reviewers for helpful discussions and comments on the manuscript. We also thank Sam Levin for his help with the `ipmr` package. Financial support for collecting the demographic data used in this study was provided by the U.S. National Science Foundation (awards INT 98-06351, DEB-0309819, DEB-0614149,DEB-0614339, and DEB-1948607). This article is publication no. \_\_\_ in the BDFFP Technical series. The authors declare no conflicts of interest.

@@

ERS contributed to the conceptualization, methodology, formal analysis, and led the writing of the original draft. EMB contributed to the conceptualization, methodology, writing, and, acquired funding.

@@

```{r #fig-lifecycle,  echo=FALSE, out.width="85%",fig.align='center'}
#| fig.cap = "Life cycle diagram of *Heliconia acuminata*. Each transition is associated with an equation for a vital rate function.  The functions shown on the diagram correspond to those used to construct a general, density-independent, deterministic IPM. The table below shows the equivalent equations for stochastic, kernel-resampled IPMs and stochastic, parameter-resampled IPMs."

```{r #fig-pop-states,  echo=FALSE, out.width="85%",fig.align='center'}
#| fig.cap = "**(A)** The proportion of _Heliconia acuminata_ populations in different size/stage classes when simulating population dyammics with three different Integral Projection Models. Results are shown for the first 250 iterations of populations in both in Continuous Forest (CF) and Forest Fragments (FF); for the criteria used to define the size categories see @tbl-plantcats. **(B)** For each interation, the relative proportion of the population in each size class (FF vs CF). Note that this excludes transient dynamics (iterations 1-30). Values on the 1-1 line indicate an iteration where CF and FF have the same proportion of the population in a given size class."
# note flowering prob for rp1 <0.25 but for rp2>0.2???'
# include _seedlings_ (a discrete category in the IPMs), _pre-reproductive 1_ (log(size) = 0–2.5, average survival < 0.9, probability of flowering \\approx 0), _pre-reproductive 2_ (log(size) = 2.5–4.5, average survival probabilty > 0.8, probability of flowering \\approx 0), _reproductive 1_ (log(size) = 4.5–6, average survival probability >0.95, flowering probability < 0.25), and _reproductive 2_ (log(size) = 6+, average survival probability >0.95, flowering probability > 0.2)
knitr::include_graphics(here(fig_pop_states), rel_path = FALSE)
```

```{r #fig-lambdas,  echo=FALSE, out.width="85%",fig.align='center'}
#| fig.cap = "Distribution of 900 values of \u03BB projected with (A) Stochastic, kernel-resampled IPMs and (B) Stochastic, paramenter-resampled IPMs. IPMs were used to project \u03BB for both Continuous Forest (green) and Forest Fragments (gray). The solid line indicates the mean value, the dashed line indicates the value of \u03BB $\\lambda$ in that habitat projected with Deterministic IPMs."

# The figures is currently generated with "plotting_lambdas.R, need to move to targets_r 
path <- here("docs/figures/lambdas_fig.png")
knitr::include_graphics(path, rel_path = FALSE)
```

```{r #fig-var,  echo=FALSE, out.width="85%",fig.align='center'}
#| fig.cap = "------"

# The figures is currently generated with "plotting_lambdas.R, need to move to targets_r 
path <- here("docs/figures/relative_variances_fig.png")
knitr::include_graphics(path, rel_path = FALSE)
```

```{r #fig-marginal,  echo=FALSE, out.width="85%",fig.align='center'}
#| fig.cap = "------"

# The figures is currently generated with "plotting_lambdas.R, need to move to targets_r 
path <- here("docs/figures/marginal_stage_fig.png")
knitr::include_graphics(path, rel_path = FALSE)
```

@@

```{r aic, results='asis'}
#| label: tbl-aic
#| tbl-cap: "Comparison of vital rate models used to build IPM.  The 'Effect of Environment' column describes how environmental effects were included in models. Those with 'none' were used to build deterministic IPMs; those with a random effect of year were used to build stochastic, kernel-resampled IPMs; and those with a distributed lag non-linear model (DLNM) were used to build stochastic, parameter-resampled IPMs. 'edf' is the estimated degrees of freedom of the penalized GAM. \u0394AIC is calculated within each habitat and vital rate combination.  \u0394AIC within 2 indicates models are equivalent."
#| 
aic_tbl_df %>%
  group_by(habitat, model) %>% 
  # mutate(rank = row_number()) %>% 
  ungroup() %>% 
  mutate(habitat = toupper(habitat),
         env_effect_incl = ifelse(
           env_effect_incl != "DLNM",
           to_sentence_case(as.character(env_effect_incl)),
           as.character(env_effect_incl)
         )
  )%>% 
  mutate(
    model = fct_relevel(
      model,
      c("surv", "size", "flwr", "surv_sdlg", "size_sdlg")
    ),
    model = fct_recode(
      model,
      "Survival"          = "surv",
      "Growth"            = "size",
      "Flowering"         = "flwr",
      "Seedling survival" = "surv_sdlg",
      "Seedling growth"   = "size_sdlg"
    )
  ) %>%

@@

  arrange(habitat, model) %>% 
  mutate(habitat=c(rep("Continuous Forest",15),rep("Forest Fragments",15))) %>% 
  mutate(model=(rep(c("Survival","","","Growth","","","Flowering","","","Seedling survival","","","Seedling growth","",""),2)))  %>%
  select(
    "Habitat" = "habitat",
    "Vital Rate" = "model",
    "Effect of Environment" = "env_effect_incl",
    #"Model" = "RHS",
    "edf" = "df",
    "\u0394AIC" = "dAIC"
  )  %>% 
  mutate(Habitat = str_replace_all(
    Habitat,
    c(
      "Continuous Forest" = "",
      "Forest Fragments" = ""
      )
  )) %>% 
  kable(digits = 2,
  format = "latex",
  align = "cllcc",
  escape = FALSE,
  row.names = FALSE,
  booktabs = T,
  linesep = ""
) %>%
  kable_styling(
    bootstrap_options = c("hover"),
    # full_width = F,
    latex_options = c("scale_down"),
    # latex_options = c("hold_position"),
    font_size = 12,
    position = "center"
  ) %>% 
# %>% 
  # collapse_rows(columns = 1:2, valign = "top")
  column_spec(1, width = "10em") %>% 
  pack_rows("Continuous Forest", 1, 15) %>%
  pack_rows("Forest Fragments", 16, 30)

# pandoc.table(
#   justify = c( "left", "left","left", "right",  "right"),
#   full_width = FALSE
#   # ,
#   # caption = Comparison of vital rate models used to build IPM.  The 'Effect of Environment' column describes how environmental effects were included in models. Those with 'none' were used to build deterministic IPMs; those with a random effect of year were used to build stochastic, kernel-resampled IPMs; and those with a distributed lag non-linear model (DLNM) were used to build stochastic, parameter-resampled IPMs. 'edf' is the estimated degrees of freedom of the penalized GAM. \u0394AIC is calculated within each habitat and vital rate combination.  \u0394AIC within 2 indicates models are equivalent."
# )

@@

lambda_table %>% 
  mutate(habitat = toupper(habitat)) %>% 
  mutate(across(c(est, lower, upper), ~format(round(.,4)))) %>% 
  mutate(lambda = glue::glue("{est} ({lower}, {upper})")) %>%
  mutate(ipm = str_replace_all(
    ipm,
    c(
      "det" = "Deterministic",
      "stoch" = "Stochastic, kernel-resampled",
      "dlnm" = "Stochastic, parameter-resampled"
    )
  )) %>% 
  select(IPM = ipm, Habitat = habitat, "$\\lambda$" = lambda) %>% 
  mutate(IPM = str_replace_all(
    IPM,
    c(
      "Deterministic" = "",
      "Stochastic, kernel-resampled" = "",
      "Stochastic, parameter-resampled" = ""
      )
  )) %>% 
  kable(digits = 2,
  format = "latex",
  align = "lcc",
  escape = FALSE,
  row.names = FALSE,
  booktabs = T,
  linesep = ""
) %>% 
  kable_styling(
    bootstrap_options = c("hover"),
    # full_width = F,
    latex_options = c("scale_down"),
    # latex_options = c("hold_position"),
    font_size = 12,
    position = "center"
  ) %>% 
  # collapse_rows(columns = 1:2, valign = "top")
  column_spec(1, width = "15em") %>% 
  pack_rows("Deterministic", 1, 2) %>%
  pack_rows("Stochastic, kernel resampled", 3, 4) %>% 
  pack_rows("Stochastic, parameter-resampled", 5, 5)

# 
# pandoc.table(
#   digits = 5,
#   keep.trailing.zeros = TRUE,
#   justify = c("left", "center", "center"),
#   full_width = FALSE
  # ,
  # caption = "(\\#tab-lambdas) Population growth rates for continuous forest (CF) and forest fragments (FF) under different kinds of IPMs with bootstrapped, bias-corrected, 95% confidence intervals."
# )

@@

```{r tbl-plantcats, results='asis'}
#| label: tbl-plantcats
#| tbl-cap: "Size and stage categories used for comparing _Heliconia acuminata_ population structure. Note that seedlings are a discrete size class not based on size (see _Methods_ for additional details)."
